---
title: "Draft Class Strength and NBA Tanking Incentives"
author: "Nico L, Leo K, Colin C"
format: docx
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(broom)
library(ggrepel)
library(scales)
library(knitr)

team_box <- if (file.exists("data/nba_team_box_2010_2026.rds")) {
  readRDS("data/nba_team_box_2010_2026.rds")
} else {
  hoopR::load_nba_team_box(seasons = 2010:2026)
}

player_box <- if (file.exists("data/nba_player_box_2010_2026.rds")) {
  readRDS("data/nba_player_box_2010_2026.rds")
} else {
  hoopR::load_nba_player_box(seasons = 2010:2026)
}

draft_prospects <- read_csv(
  "data/crafted_prospects_2010_2026.csv",
  show_col_types = FALSE
)

draft_strength <- read_csv(
  "data/crafted_draft_strength_2010_2026.csv",
  show_col_types = FALSE
)

theme_set(theme_minimal(base_size = 12))

collapse_repeated_rows <- function(data, columns) {
  for (column in columns) {
    values <- as.character(data[[column]])
    repeated <- values == lag(values, default = NA_character_)
    values[replace_na(repeated, FALSE)] <- ""
    data[[column]] <- values
  }

  data
}
```

```{r data-build}
nba_teams <- c(
  "ATL", "BOS", "BKN", "NJ", "CHA", "CHI", "CLE", "DAL", "DEN", "DET",
  "GS", "HOU", "IND", "LAC", "LAL", "MEM", "MIA", "MIL", "MIN", "NO",
  "NY", "OKC", "ORL", "PHI", "PHX", "POR", "SA", "SAC", "TOR", "UTAH",
  "WSH"
)

conference_lookup <- tibble(
  team_abbreviation = nba_teams,
  conference = c(
    "East", "East", "East", "East", "East", "East", "East", "West", "West", "East",
    "West", "West", "East", "West", "West", "West", "East", "East", "West", "West",
    "East", "West", "East", "East", "West", "West", "West", "West", "East", "West",
    "East"
  )
)

position_group_from_label <- function(position, archetype = rep(NA_character_, length(position))) {
  case_when(
    position %in% c("PG", "SG", "G") ~ "Guard",
    position %in% c("SF", "F") ~ "Wing",
    position %in% c("PF", "C") ~ "Big",
    archetype == "Guard" ~ "Guard",
    archetype %in% c("Wing", "Forward") ~ "Wing",
    archetype == "Big" ~ "Big",
    TRUE ~ NA_character_
  )
}

regular_games <- team_box |>
  filter(
    season_type == 2,
    team_abbreviation %in% nba_teams,
    season %in% 2010:2026
  ) |>
  mutate(
    win = team_score > opponent_team_score,
    cutoff_date = as.Date(paste0(season, "-03-01")),
    season_period = if_else(game_date < cutoff_date, "pre_march", "late")
  ) |>
  left_join(conference_lookup, by = "team_abbreviation")

team_periods <- regular_games |>
  group_by(season, team_abbreviation, conference, season_period) |>
  summarise(
    games = n(),
    wins = sum(win, na.rm = TRUE),
    losses = games - wins,
    point_diff = mean(team_score - opponent_team_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = season_period,
    values_from = c(games, wins, losses, point_diff),
    values_fill = 0
  )

compute_lottery_odds <- function(weights, draws) {
  seeds <- seq_along(weights)
  top_pick_odds <- numeric(length(weights))
  top4_pick_odds <- numeric(length(weights))

  add_draw_path <- function(remaining, drawn, path_probability) {
    if (length(drawn) == draws) {
      final_order <- c(drawn, setdiff(seeds, drawn))
      top_pick_odds[final_order[1]] <<- top_pick_odds[final_order[1]] + path_probability
      top4_pick_odds[final_order[1:4]] <<- top4_pick_odds[final_order[1:4]] + path_probability
      return(invisible(NULL))
    }

    total_weight <- sum(weights[remaining])
    for (seed in remaining) {
      add_draw_path(
        remaining = setdiff(remaining, seed),
        drawn = c(drawn, seed),
        path_probability = path_probability * weights[seed] / total_weight
      )
    }
  }

  add_draw_path(seeds, integer(), 1)

  tibble(
    lottery_seed_march = seeds,
    top_pick_odds = top_pick_odds,
    top4_pick_odds = top4_pick_odds
  )
}

lottery_odds <- bind_rows(
  compute_lottery_odds(
    weights = c(250, 199, 156, 119, 88, 63, 43, 28, 17, 11, 8, 7, 6, 5),
    draws = 3
  ) |>
    mutate(post_reform = FALSE),
  compute_lottery_odds(
    weights = c(140, 140, 140, 125, 105, 90, 75, 60, 45, 30, 20, 15, 10, 5),
    draws = 4
  ) |>
    mutate(post_reform = TRUE)
) |>
  group_by(post_reform) |>
  arrange(lottery_seed_march, .by_group = TRUE) |>
  mutate(
    marginal_top4_pick_gain = pmax(0, lag(top4_pick_odds) - top4_pick_odds),
    marginal_top4_pick_gain = replace_na(marginal_top4_pick_gain, 0),
    marginal_top4_pick_gain = if_else(marginal_top4_pick_gain < 1e-8, 0, marginal_top4_pick_gain)
  ) |>
  ungroup()

team_data <- team_periods |>
  mutate(
    pre_win_pct = wins_pre_march / games_pre_march,
    postseason_cutline = if_else(season >= 2021, 10L, 8L)
  ) |>
  group_by(season, conference) |>
  arrange(desc(pre_win_pct), desc(wins_pre_march), .by_group = TRUE) |>
  mutate(
    conference_rank_march = row_number(),
    cutline_wins = wins_pre_march[postseason_cutline[1]],
    cutline_losses = losses_pre_march[postseason_cutline[1]],
    games_back_cutline = pmax(
      0,
      ((cutline_wins - wins_pre_march) + (losses_pre_march - cutline_losses)) / 2
    )
  ) |>
  ungroup() |>
  group_by(season) |>
  arrange(pre_win_pct, wins_pre_march, point_diff_pre_march, .by_group = TRUE) |>
  mutate(
    lottery_position_march = row_number(),
    lottery_seed_march = pmin(lottery_position_march, 14L)
  ) |>
  ungroup() |>
  mutate(
    late_win_pct = wins_late / games_late,
    expected_late_wins = games_late * pre_win_pct,
    expected_sd = sqrt(pmax(games_late * pre_win_pct * (1 - pre_win_pct), 0.001)),
    late_win_shortfall = expected_late_wins - wins_late,
    tanking_index = late_win_shortfall / expected_sd,
    outside_postseason_march = conference_rank_march > postseason_cutline,
    likely_lottery_candidate = outside_postseason_march &
      (games_back_cutline >= 3 | pre_win_pct < 0.400),
    post_reform = season >= 2019,
    season_2012 = season == 2012,
    season_2020 = season == 2020,
    season_2021 = season == 2021,
    disrupted_season = season %in% c(2012, 2020, 2021),
    lottery_era = if_else(post_reform, "Post-reform (2019-2026)", "Pre-reform (2010-2018)")
  ) |>
  left_join(
    draft_strength |>
      select(
        draft_year,
        avg_rating,
        elite_count,
        above_avg_count,
        top_prospect_rating,
        top3_avg_rating,
        top4_avg_rating,
        top5_avg_rating
      ),
    by = c("season" = "draft_year")
  ) |>
  left_join(
    lottery_odds,
    by = c("post_reform", "lottery_seed_march")
  ) |>
  mutate(
    marginal_lottery_incentive = top4_avg_rating * marginal_top4_pick_gain,
    marginal_lottery_incentive_top35 = avg_rating * marginal_top4_pick_gain
  )

team_position_need <- player_box |>
  filter(
    season_type == 2,
    season %in% 2010:2026,
    team_abbreviation %in% nba_teams
  ) |>
  mutate(
    cutoff_date = as.Date(paste0(season, "-03-01")),
    position_group = position_group_from_label(athlete_position_abbreviation),
    minutes_num = suppressWarnings(as.numeric(minutes)),
    box_value = points + rebounds + assists + steals + blocks - turnovers
  ) |>
  filter(
    game_date < cutoff_date,
    !is.na(position_group),
    !is.na(minutes_num),
    minutes_num > 0
  ) |>
  group_by(season, team_abbreviation, position_group) |>
  summarise(
    minutes = sum(minutes_num, na.rm = TRUE),
    box_value = sum(box_value, na.rm = TRUE),
    value_per36 = 36 * box_value / minutes,
    .groups = "drop"
  ) |>
  group_by(season, position_group) |>
  mutate(
    position_value_z = as.numeric(scale(value_per36)),
    position_need = pmax(0, -position_value_z)
  ) |>
  ungroup()

draft_position_supply <- draft_prospects |>
  mutate(
    position_group = position_group_from_label(position, archetype),
    rating = as.numeric(rating)
  ) |>
  filter(
    draft_year %in% 2010:2026,
    !is.na(position_group),
    !is.na(rating)
  ) |>
  arrange(draft_year, desc(rating)) |>
  group_by(draft_year) |>
  slice_head(n = 35) |>
  ungroup() |>
  mutate(draft_value = pmax(rating - 50, 0)) |>
  group_by(draft_year, position_group) |>
  summarise(
    draft_value = sum(draft_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  complete(
    draft_year,
    position_group = c("Guard", "Wing", "Big"),
    fill = list(draft_value = 0)
  ) |>
  group_by(draft_year) |>
  mutate(
    total_draft_value = sum(draft_value, na.rm = TRUE),
    draft_value_share = if (total_draft_value[1] > 0) {
      draft_value / total_draft_value[1]
    } else {
      rep(1 / 3, n())
    }
  ) |>
  ungroup()

team_fit_scores <- team_position_need |>
  left_join(
    draft_position_supply,
    by = c("season" = "draft_year", "position_group")
  ) |>
  mutate(draft_value_share = replace_na(draft_value_share, 0)) |>
  group_by(season, team_abbreviation) |>
  summarise(
    team_fit_score = sum(position_need * draft_value_share, na.rm = TRUE),
    strongest_need = position_group[which.max(position_need)],
    .groups = "drop"
  )

player_usage <- player_box |>
  filter(
    season_type == 2,
    season %in% 2010:2026,
    team_abbreviation %in% nba_teams
  ) |>
  mutate(
    cutoff_date = as.Date(paste0(season, "-03-01")),
    season_period = if_else(game_date < cutoff_date, "pre_march", "late"),
    minutes_num = suppressWarnings(as.numeric(minutes)),
    minutes_played = replace_na(minutes_num, 0),
    starter_flag = starter %in% TRUE
  )

team_games <- regular_games |>
  distinct(season, team_abbreviation, game_id, game_date, season_period)

late_core_appearances <- player_usage |>
  filter(season_period == "late", minutes_played > 0) |>
  distinct(season, team_abbreviation, athlete_id)

core_players <- player_usage |>
  filter(season_period == "pre_march", minutes_played > 0) |>
  group_by(season, team_abbreviation, athlete_id, athlete_display_name) |>
  summarise(
    pre_starts = sum(starter_flag, na.rm = TRUE),
    pre_minutes = sum(minutes_played, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(season, team_abbreviation) |>
  arrange(desc(pre_starts), desc(pre_minutes), .by_group = TRUE) |>
  slice_head(n = 3) |>
  ungroup() |>
  semi_join(
    late_core_appearances,
    by = c("season", "team_abbreviation", "athlete_id")
  )

core_usage <- core_players |>
  select(season, team_abbreviation, athlete_id) |>
  inner_join(
    team_games,
    by = c("season", "team_abbreviation"),
    relationship = "many-to-many"
  ) |>
  left_join(
    player_usage |>
      select(season, team_abbreviation, game_id, athlete_id, minutes_played),
    by = c("season", "team_abbreviation", "game_id", "athlete_id")
  ) |>
  mutate(minutes_played = replace_na(minutes_played, 0)) |>
  group_by(season, team_abbreviation, season_period) |>
  summarise(
    core_minutes_per_game = sum(minutes_played, na.rm = TRUE) / n_distinct(game_id),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = season_period,
    values_from = core_minutes_per_game,
    names_prefix = "core_mpg_"
  ) |>
  mutate(
    core_usage_drop = core_mpg_pre_march - core_mpg_late,
    core_usage_drop_rate = if_else(
      core_mpg_pre_march > 0,
      core_usage_drop / core_mpg_pre_march,
      NA_real_
    )
  )

team_data <- team_data |>
  left_join(team_fit_scores, by = c("season", "team_abbreviation")) |>
  left_join(core_usage, by = c("season", "team_abbreviation"))
```

## Abstract

This project studies whether stronger NBA draft classes are associated with more late-season underperformance among teams with weak playoff outlooks. Using NBA team and player game logs from 2010 through 2026 and CraftedNBA draft prospect ratings, we build a team-season dataset that compares each team's performance before March 1 to its performance after March 1. As the project developed, we moved away from a simple binary definition of tanking and instead used a standardized `tanking_index`, which measures how many fewer games a team won late in the season than expected from its pre-March record. We also changed the draft-strength measure from a broad top-35 average to the average rating of the top four prospects, because lottery incentives are most directly tied to access to the top of the draft. The main win-loss result is suggestive rather than conclusive: likely lottery teams with larger position-adjusted draft incentives tend to have larger late-season win shortfalls, but the evidence is limited by the fact that draft quality varies mostly by season and tanking intent is not directly observable. A fourth exploratory analysis finds a stronger relationship between top-four draft strength and reduced late-season minutes for continuing core players, which points to roster management as a useful follow-up tanking signal.

## I. Introduction

This project asks: **Are stronger NBA draft classes associated with more late-season underperformance among teams with weak playoff outlooks, especially when a team's lottery position gives it more to gain from losing?**

The original project proposal focused broadly on whether stronger draft classes create stronger incentives for NBA teams to tank. As we looked more closely at the data, the question became more specific in three ways. First, we stopped treating tanking as something we could directly observe. Instead, we measure late-season underperformance among teams with weak postseason outlooks. Second, we changed the draft-strength measure to focus on the top four prospects rather than the average quality of a much larger prospect group, since the NBA lottery primarily gives teams access to the very top of the draft. Third, we made the incentive team-specific by combining top-four draft strength with each team's estimated lottery position entering March. This matters because the same draft class does not create the same incentive for every team.

This framing gives us a more careful final research question: **among likely lottery teams, are stronger top-end draft classes associated with larger late-season win shortfalls after accounting for each team's lottery position and pre-March strength?** The answer matters because tanking debates often focus on lottery odds alone, while the value of losing should also depend on the players available in that draft.

## II. Background

This question is about incentives, but our data cannot directly observe intent. We therefore measure **tanking signals**, not confirmed tanking. A tanking signal is a team-season where a team had a weak postseason outlook entering March and then performed worse than expected late in the regular season. This framing is important because a team can lose late-season games for many reasons, including injuries, player development, fatigue, roster trades, or strategic resting.

The NBA draft lottery determines which non-playoff teams receive the first picks in the draft. Worse teams generally receive better lottery odds, which creates the basic incentive problem: a team that is unlikely to make the playoffs may benefit from losing more late-season games if doing so improves its draft position. The [NBA changed its lottery odds beginning with the 2019 draft](https://www.nba.com/news/nba-board-governors-approves-changes-draft-lottery-system), flattening the odds at the top. Under the [current format](https://www.nba.com/news/nba-draft-lottery-explainer?lv=true), the three worst teams each have a 14.0% chance at the first pick, while other lottery teams receive lower odds. The modern lottery determines the first four picks, which is why our preferred draft-strength measure focuses on the average quality of the top four prospects rather than the full draft class.

Prior research has studied tanking as a response to draft lottery incentives. [Taylor and Trogdon (2002)](https://ideas.repec.org/a/ucp/jlabec/v20y2002i1p23-41.html) and [Price et al. (2010)](https://journals.sagepub.com/doi/10.1177/1527002510363103) examine whether NBA teams lose more often when the reward for losing is larger. Our project builds on that idea by adding a second source of incentive: the quality of the incoming draft class. Instead of treating every draft as equally valuable, we ask whether likely lottery teams appear to underperform more when the player pool at the top of the draft is stronger and when moving down in the standings would increase their chance of landing a top-four pick.

## III. Data Gathering

We use two main source families. First, NBA game logs come from the [`hoopR`](https://hoopr.sportsdataverse.org/) package. The team game logs provide the game date, team, opponent, score, and win/loss result. We also use player game logs from `hoopR` to build an exploratory team-fit measure based on pre-March production by position group. Second, draft class quality comes from [CraftedNBA draft class rankings](https://craftednba.com/draft/class-rankings) and prospect-level data from CraftedNBA's draft pages. The main draft-quality variable is `top4_avg_rating`, the average CraftedNBA rating of the top four prospects in that draft class. This is more closely tied to the lottery mechanism than the average of a larger prospect group because lottery teams are mostly trying to access the very top of the draft. We also keep the top-35 average rating, prospect positions, number of elite prospects, number of above-average prospects, and rating of the top prospect as alternate draft-strength information.

The final dataset contains `r nrow(team_data)` team-seasons from `r min(team_data$season)` through `r max(team_data$season)`. Each row is one team in one season. Raw draft quality changes only at the season level, so the analysis has `r n_distinct(team_data$season)` draft-class observations, not `r nrow(team_data)` independent draft-strength observations. To make the main incentive variable vary by team, we combine draft strength with each team's estimated March 1 lottery position and probability of landing a top-four pick. This does not fully eliminate the season-level limitation, but it better represents the fact that the same draft class creates different incentives for the worst team, the fifth-worst team, and a fringe lottery team.

We split each season at March 1. The pre-March portion gives a baseline estimate of team strength before the final stretch of the regular season. The March-onward portion is the window where late-season incentives are more likely to matter. The main preprocessing steps were filtering to regular-season games, aggregating team records before and after March 1, ranking teams by pre-March league record to estimate lottery position, merging in draft class ratings, and constructing the late-season underperformance measures described below.

The lottery-position adjustment uses published NBA lottery weights. Before the 2019 reform, the lottery drew the first three picks; after the reform, it draws the first four picks. We use those weights to calculate each seed's probability of finishing in the top four, then calculate how much that probability would rise from moving one lottery slot worse. We use standard lottery odds by seed rather than final lottery odds after end-of-season tiebreakers because the model is based on information available around March 1.

Key constructed variables:

-   `pre_win_pct` and `late_win_pct` are each team's winning percentage before March 1 and on or after March 1.

    $$
    \text{Pre-March win pct} = \frac{W_{\text{pre}}}{G_{\text{pre}}},
    \qquad
    \text{Late win pct} = \frac{W_{\text{late}}}{G_{\text{late}}}
    $$

    `pre_win_pct` is used as a baseline team-strength control because a team that was already very weak before March should not be evaluated the same way as a team that was close to .500.

-   `games_back_cutline` measures how far a team was from the postseason/play-in cutoff entering March. For seasons before the play-in era, the cutoff is the 8th seed. For seasons beginning in 2021, the cutoff is the 10th seed. We use this to help identify teams with weak playoff outlooks.

    $$
    \text{Games back} =
    \max\left(
      0,
      \frac{(W_{\text{cutline}} - W_{\text{team}})
      + (L_{\text{team}} - L_{\text{cutline}})}{2}
    \right)
    $$

-   `likely_lottery_candidate` is our sample filter for teams that plausibly had an incentive to prioritize draft position over late-season wins.

    $$
    \text{Likely lottery candidate} =
    \begin{cases}
    1, & \text{outside cutoff and }(\text{games back} \geq 3
    \text{ or pre-March win pct} < 0.400)\\
    0, & \text{otherwise}
    \end{cases}
    $$

    This is not an outcome variable; it defines the group of teams used in the formal models.

-   `lottery_seed_march` is an estimated lottery position entering March, based on league-wide pre-March record. The worst pre-March record is seed 1, the second-worst is seed 2, and so on through the 14-team lottery. This is an approximation because final lottery order depends on end-of-season records and tiebreakers.

-   `top4_avg_rating` is the average CraftedNBA rating of the top four prospects in a draft class.

    $$
    \text{Top-four average rating} =
    \frac{r_1 + r_2 + r_3 + r_4}{4}
    $$

    Higher values mean a stronger top of the draft. This variable is the same for every team in the same season, so raw top-four draft-strength models are interpreted as draft-class-level evidence.

-   `marginal_lottery_incentive` is our team-specific draft incentive. First, we calculate how much a team's probability of landing a top-four pick would increase if it moved one lottery slot worse. Then we multiply that probability gain by `top4_avg_rating`.

    $$
    \text{Marginal lottery incentive} =
    \text{Top-four average rating}
    \times
    \text{Marginal top-four-pick probability gain}
    $$

    For example, if moving from the fifth-worst to fourth-worst record increases a team's top-four-pick probability, that gain is more valuable in a year with a stronger top four. Teams with no added top-four-pick probability from moving one slot worse, such as the worst team, receive a marginal incentive of 0.

-   `tanking_index` is the main win-loss outcome. It compares a team's actual March-onward wins to the number of late-season wins expected from its pre-March winning percentage.

    $$
    \widehat{W}_{\text{late}} =
    G_{\text{late}} \times p_{\text{pre}}
    $$

    $$
    \text{Tanking index} =
    \frac{\widehat{W}_{\text{late}} - W_{\text{late}}}
    {\sqrt{G_{\text{late}}p_{\text{pre}}(1-p_{\text{pre}})}}
    $$

    Here, $p_{\text{pre}}$ is the team's pre-March winning percentage. The code uses a small lower bound on the denominator for numerical stability.

    Positive values mean the team won fewer late-season games than expected. Values near 0 mean it performed about as expected. Negative values mean it overperformed late in the season.

-   `team_fit_score` is an exploratory team-specific measure of whether a team's weakest position groups match the strongest position groups in that year's draft. We group players and prospects into three position groups: guards, wings, and bigs. For each team-position group before March 1, we calculate a simple production rate:

    $$
    \text{Value per 36} =
    36 \times
    \frac{\text{points} + \text{rebounds} + \text{assists}
    + \text{steals} + \text{blocks} - \text{turnovers}}
    {\text{minutes}}
    $$

    Within each season and position group, we standardize value per 36 across teams. A team gets a positive need score only where it is below average:

    $$
    \text{Position need}_{g} =
    \max\left(0,-z(\text{Value per 36}_{g})\right)
    $$

    We then measure draft supply using the top 35 CraftedNBA prospects. Each prospect's draft value is:

    $$
    \text{Draft value}_{j} =
    \max(\text{rating}_{j} - 50, 0)
    $$

    For each position group, draft value is converted into a share of all top-35 draft value:

    $$
    \text{Draft value share}_{g} =
    \frac{\sum_{j \in g}\text{Draft value}_{j}}
    {\sum_{j}\text{Draft value}_{j}}
    $$

    The final score is:

    $$
    \text{Team fit score} =
    \sum_{g \in \{\text{Guard, Wing, Big}\}}
    \text{Position need}_{g}
    \times
    \text{Draft value share}_{g}
    $$

    Higher values mean a team was weak at positions where that draft had more high-rated prospects. This is a rough proxy for roster fit, not a team-specific draft board.

-   `core_usage_drop_rate` is an exploratory roster-management outcome. For each team-season, we identify the team's top three pre-March core players by starts, breaking ties by minutes. We keep only core players who appeared for the same team after March 1, so traded players are not counted as "benched." We then calculate their combined minutes per team game before and after March:

    $$
    \text{Core usage drop rate} =
    \frac{\text{Pre-March core MPG} - \text{Late core MPG}}
    {\text{Pre-March core MPG}}
    $$

    Positive values mean the team's main continuing pre-March players were used less after March 1. This does not prove healthy starters were benched; injuries, rest, development minutes, and coaching decisions can also reduce minutes.

-   `season_2012`, `season_2020`, and `season_2021` are separate controls for disrupted seasons. We use separate indicators because the 2012 lockout season, 2020 bubble season, and 2021 shortened COVID-era season may have affected late-season performance in different ways.

The `tanking_index` is the main outcome because it uses more information than a simple cutoff. For example, the 2021 Rockets were 11-21 before March, a .344 win percentage. If they had kept winning at that rate over their 40 March-onward games, they would have been expected to win about 13.8 games. They won 6, leaving a shortfall of 7.8 wins and a tanking index of 2.58. This does not prove intent, but it identifies a large late-season underperformance relative to that team's own earlier baseline. Similarly, if a team's top three continuing core players averaged 95 combined minutes per game before March and 76 after March, its core usage drop rate would be:

$$
\frac{95 - 76}{95} = 0.20
$$

or a 20% drop in core-player usage.

## IV. Data Analysis

This section tells the main descriptive story in the data before moving to regression models. The key questions are whether likely lottery teams look different from other teams, whether late-season underperformance is rare or common, and whether stronger top-end draft classes line up with larger late-season win shortfalls.

```{r summary-table}
summary_table <- team_data |>
  mutate(group = if_else(likely_lottery_candidate, "Likely lottery candidates", "Other teams")) |>
  group_by(lottery_era, group) |>
  summarise(
    N = n(),
    `Pre win pct` = mean(pre_win_pct, na.rm = TRUE),
    `Late win pct` = mean(late_win_pct, na.rm = TRUE),
    `Tanking index` = mean(tanking_index, na.rm = TRUE),
    .groups = "drop"
  ) |>
  transmute(
    Era = lottery_era,
    Group = group,
    N,
    `Pre win pct`,
    `Late win pct`,
    `Tanking index`
  )

kable(
  collapse_repeated_rows(summary_table, "Era"),
  digits = 3,
  caption = "Table 1. Summary statistics by lottery era and team group."
)
```

Table 1 shows that likely lottery candidates are much weaker than other teams before March, as expected. The average win percentage change from pre-March to late season is small in both eras, which suggests that most weak teams do not suddenly collapse late in the year. The final regression relies on the full `tanking_index`, which measures degree of late-season underperformance instead of forcing teams into a yes/no collapse category.

```{r visual-expected-vs-actual, fig.width=8, fig.height=5, fig.cap="Figure 1. Late-season performance compared with pre-March performance. Each point is one team-season from 2010 through 2026. The dashed line shows equal pre-March and March-onward win percentages. Triangles are likely lottery candidates. Color shows the tanking index, where higher values mean larger late-season win shortfalls. Labels identify the largest tanking-index seasons among likely lottery candidates."}
label_data <- team_data |>
  filter(likely_lottery_candidate) |>
  slice_max(tanking_index, n = 12, with_ties = FALSE)

ggplot(team_data, aes(x = pre_win_pct, y = late_win_pct)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(
    aes(color = tanking_index, shape = likely_lottery_candidate),
    size = 2.4,
    alpha = 0.72
  ) +
  geom_text_repel(
    data = label_data,
    aes(label = paste(team_abbreviation, season)),
    size = 2.8,
    seed = 470,
    box.padding = 0.35,
    min.segment.length = 0
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_color_gradient2(
    low = "#355C7D",
    mid = "gray75",
    high = "#C0392B",
    midpoint = 0
  ) +
  labs(
    title = "Late-season performance compared with pre-March performance",
    x = "Pre-March win percentage",
    y = "March-onward win percentage",
    color = "Tanking index",
    shape = "Likely lottery candidate"
  )
```

Figure 1 shows that most teams cluster near the diagonal, meaning their late-season winning percentage was similar to their pre-March winning percentage. The likely lottery candidates mostly appear in the lower-left portion of the graph, which confirms that the candidate definition is identifying weaker teams. The darker red points are teams with larger positive tanking-index values: they won substantially fewer late-season games than expected from their earlier record. This is why the regression below uses the full `tanking_index` rather than a binary collapse label.

```{r visual-draft-strength, fig.width=8, fig.height=5, fig.cap="Figure 2. Top-four draft strength and average late-season underperformance among likely lottery candidates. Each point is one NBA season. The x-axis is the CraftedNBA average rating of the top four prospects, the y-axis is the mean tanking index among likely lottery candidates, and point size is the number of likely lottery team-seasons in that year. Fitted lines are descriptive and are shown separately for the pre-reform and post-reform lottery eras."}
season_summary <- team_data |>
  filter(likely_lottery_candidate) |>
  group_by(season, lottery_era) |>
  summarise(
    top4_avg_rating = first(top4_avg_rating),
    mean_tanking_index = mean(tanking_index, na.rm = TRUE),
    candidate_count = n(),
    .groups = "drop"
  )

ggplot(season_summary, aes(x = top4_avg_rating, y = mean_tanking_index, color = lottery_era)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_point(aes(size = candidate_count), alpha = 0.85) +
  geom_text_repel(aes(label = season), size = 2.9, seed = 470, min.segment.length = 0) +
  scale_color_manual(
    values = c(
      "Pre-reform (2010-2018)" = "#5D6D7E",
      "Post-reform (2019-2026)" = "#B23A48"
    )
  ) +
  scale_size_continuous(
    range = c(2.8, 7.5),
    breaks = pretty_breaks(n = 4),
    labels = \(x) paste(x, "teams"),
    name = "Likely lottery teams\nin season"
  ) +
  labs(
    title = "Top-four draft strength and late-season underperformance",
    x = "CraftedNBA average top-four prospect rating",
    y = "Mean tanking index among likely lottery candidates",
    color = "Lottery era"
  )
```

Figure 2 provides exploratory context rather than the final model specification. The pre-reform and post-reform fitted lines are shown descriptively, but the final regression below pools all years together. The graph still shows why the conclusion should be cautious: there are only 17 season-level draft class observations, and 2021 remains unusual because it was affected by COVID-era scheduling and roster disruptions.

Although the model uses team-season observations, Figure 2 summarizes outcomes at the season level because draft strength is constant for all teams in a given draft year. Figure 3 shows the same draft-strength relationship using all likely lottery team-seasons.

```{r visual-team-season-observations, fig.width=8, fig.height=5.4, fig.cap="Figure 3. Team-season observations for top-four draft strength and late-season underperformance among likely lottery candidates. Each point is one likely lottery team-season. The x-axis is jittered slightly because every team in the same season has the same top-four draft rating. Color shows season, shape shows lottery era, and the black diamonds show the season-level averages from Figure 2."}
candidate_plot_data <- team_data |>
  filter(likely_lottery_candidate) |>
  mutate(label = if_else(tanking_index >= quantile(tanking_index, 0.94, na.rm = TRUE),
                         paste(team_abbreviation, season),
                         NA_character_))

ggplot(candidate_plot_data, aes(x = top4_avg_rating, y = tanking_index)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray55") +
  geom_jitter(
    aes(color = season, shape = lottery_era),
    width = 0.12,
    height = 0,
    size = 2.1,
    alpha = 0.68
  ) +
  geom_point(
    data = season_summary,
    aes(x = top4_avg_rating, y = mean_tanking_index),
    inherit.aes = FALSE,
    shape = 23,
    fill = "black",
    color = "white",
    size = 3.2,
    alpha = 0.9
  ) +
  geom_text_repel(
    aes(label = label),
    na.rm = TRUE,
    size = 2.7,
    seed = 470,
    box.padding = 0.35,
    min.segment.length = 0
  ) +
  scale_color_viridis_c(option = "plasma", breaks = seq(2010, 2026, by = 4)) +
  labs(
    title = "Team-season tanking index observations by top-four draft strength",
    x = "CraftedNBA average top-four prospect rating",
    y = "Tanking index",
    color = "Season",
    shape = "Lottery era"
  )
```

Figure 3 makes the observational unit clearer. Each point is a team-season, while the black diamonds are the season averages used in Figure 2. The vertical stacks show why the season-level summary can hide variation: teams in the same draft year share the same top-four draft rating, but their tanking-index values differ widely.

```{r visual-draft-rating-comparison, fig.width=8, fig.height=5.4, fig.cap="Figure 4. Comparing broad and top-four draft-strength measures. Each point is one NBA draft class from 2010 through 2026. The x-axis is the top-35 average rating, the y-axis is the top-four average rating, and color shows the mean tanking index among likely lottery candidates in that season."}
draft_measure_summary <- team_data |>
  filter(likely_lottery_candidate) |>
  group_by(season, lottery_era) |>
  summarise(
    avg_rating = first(avg_rating),
    top4_avg_rating = first(top4_avg_rating),
    mean_tanking_index = mean(tanking_index, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(draft_measure_summary, aes(x = avg_rating, y = top4_avg_rating)) +
  geom_point(aes(color = mean_tanking_index), size = 3, alpha = 0.85) +
  geom_text_repel(aes(label = season), size = 2.7, seed = 470, min.segment.length = 0) +
  scale_color_gradient2(
    low = "#5D6D7E",
    mid = "gray75",
    high = "#B23A48",
    midpoint = 0,
    name = "Mean tanking index"
  ) +
  labs(
    title = "Top-four draft strength is related to, but distinct from, broad class strength",
    x = "Top-35 average draft rating",
    y = "Top-four average draft rating"
  )
```

Figure 4 shows why the top-four measure is useful. It is related to the broader top-35 class rating, but it is not identical. The top-four measure focuses on the players lottery teams are most realistically trying to access, so it is more directly tied to the incentive mechanism than the average quality of the full top 35.

```{r visual-position-adjusted-incentive, fig.width=8, fig.height=5.4, fig.cap="Figure 5. Position-adjusted draft incentive and tanking index among likely lottery candidates. Each point is one team-season. The x-axis combines top-four draft class rating with the marginal gain in top-four-pick probability from moving one lottery slot worse. Zero-incentive points are nudged slightly to the right only to reveal overlapping observations; the fitted line and regression models use the exact, unnudged incentive values."}
figure5_data <- team_data |>
  filter(likely_lottery_candidate) |>
  arrange(season, team_abbreviation) |>
  group_by(zero_incentive = marginal_lottery_incentive == 0) |>
  mutate(
    marginal_lottery_incentive_plot = if_else(
      zero_incentive,
      marginal_lottery_incentive + ((row_number() - 1) %% 7) * 0.025,
      marginal_lottery_incentive
    )
  ) |>
  ungroup()

ggplot(figure5_data, aes(x = marginal_lottery_incentive, y = tanking_index, color = lottery_seed_march)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(
    aes(x = marginal_lottery_incentive_plot),
    alpha = 0.72,
    size = 2.3
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "gray35", linewidth = 0.8) +
  scale_color_viridis_c(
    option = "viridis",
    direction = -1,
    breaks = c(1, 4, 7, 10, 14)
  ) +
  labs(
    title = "Team-specific incentive by lottery position",
    x = "Top-four rating x marginal top-four-pick probability gain",
    y = "Tanking index",
    color = "March lottery seed"
  )
```

Figure 5 directly addresses the concern that a draft class has the same top-four rating for every team in the same year. The raw top-four draft rating is still season-level, but the position-adjusted incentive varies by team because different lottery positions have different marginal rewards for moving down in the standings. Observations at zero are meaningful: they are teams that would not gain additional top-four-pick probability by moving one slot worse, such as the worst team. We nudge the plotted zero points slightly so those overlapping observations are visible, but the model still uses the exact values. The relationship is noisier than the season-average figures, which is expected: the outcome is still affected by injuries, trades, resting, and random late-season variation.

## V. Formal Analysis

The formal analysis uses likely lottery candidates only, because those are the teams with the clearest incentive to lose late-season games. To avoid making the results depend on an arbitrary 10 percentage point cutoff, the outcome is the full `tanking_index`. This is a standardized late-season win shortfall: positive values mean a team won fewer March-onward games than expected from its pre-March record, while negative values mean it overperformed late.

All formal regression models in this section use **team-season outcomes**, not one observation per season. In other words, a team like Houston in 2021 and Orlando in 2021 are separate rows because they had different pre-March records, lottery positions, marginal lottery incentives, team-fit scores, core-player usage patterns, and late-season outcomes. However, not every explanatory variable varies at the team-season level. Raw `top4_avg_rating` is the same for every team in the same draft year, so models using only raw draft strength should be interpreted as evidence about season-level draft-class strength, not as 172 fully independent draft-strength observations. To address this, we emphasize team-specific variables such as `marginal_lottery_incentive` and `team_fit_score` when the research question is about team-level incentives, and we report season-clustered inference for the core-player usage checks that use raw top-four draft rating.

The final model's key coefficient is the standardized position-adjusted draft incentive. It asks whether teams with larger marginal lottery rewards in stronger top-four draft classes have larger tanking signals after controlling for pre-March team strength and separate indicators for the disrupted 2012, 2020, and 2021 seasons. We also include follow-up checks for whether the relationship changed after the 2019 lottery reform, whether a team's roster fit with the draft class adds explanatory power, and whether stronger top-four draft classes are associated with late-season reductions in core-player minutes.

### Analysis 1: Pre/Post Reform Interaction

```{r models}
candidate_data <- team_data |>
  filter(likely_lottery_candidate)

model_tanking_index_interaction <- lm(
  tanking_index ~ scale(top4_avg_rating) * post_reform +
    pre_win_pct +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_tanking_index <- lm(
  tanking_index ~ scale(marginal_lottery_incentive) +
    pre_win_pct +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_tanking_index_top35 <- lm(
  tanking_index ~ scale(marginal_lottery_incentive_top35) +
    pre_win_pct +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

candidate_data_no_2021 <- candidate_data |>
  filter(season != 2021)

model_tanking_index_no_2021 <- lm(
  tanking_index ~ scale(marginal_lottery_incentive) +
    pre_win_pct +
    season_2012 +
    season_2020,
  data = candidate_data_no_2021
)

model_tanking_index_fit <- lm(
  tanking_index ~ scale(top4_avg_rating) * scale(team_fit_score) +
    pre_win_pct +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage <- lm(
  core_usage_drop_rate ~ scale(top4_avg_rating) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage_interaction <- lm(
  core_usage_drop_rate ~ scale(top4_avg_rating) * post_reform +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage_position_adjusted <- lm(
  core_usage_drop_rate ~ scale(marginal_lottery_incentive) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage_position_adjusted_no_2021 <- lm(
  core_usage_drop_rate ~ scale(marginal_lottery_incentive) +
    core_mpg_pre_march +
    season_2012 +
    season_2020,
  data = candidate_data_no_2021
)

model_core_usage_no_2021 <- lm(
  core_usage_drop_rate ~ scale(top4_avg_rating) +
    core_mpg_pre_march +
    season_2012 +
    season_2020,
  data = candidate_data_no_2021
)

model_core_usage_fit <- lm(
  core_usage_drop_rate ~ scale(top4_avg_rating) * scale(team_fit_score) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage_top35 <- lm(
  core_usage_drop_rate ~ scale(avg_rating) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

cluster_vcov <- function(model, cluster) {
  X <- model.matrix(model)
  u <- resid(model)
  model_rows <- suppressWarnings(as.integer(rownames(model.frame(model))))

  if (all(!is.na(model_rows)) && max(model_rows) <= length(cluster)) {
    cluster <- cluster[model_rows]
  } else {
    cluster <- cluster[seq_along(u)]
  }

  xtx_inv <- solve(crossprod(X))
  meat <- matrix(0, ncol(X), ncol(X))

  for (cluster_id in unique(cluster)) {
    idx <- cluster == cluster_id
    X_cluster <- X[idx, , drop = FALSE]
    u_cluster <- u[idx]
    xu <- crossprod(X_cluster, u_cluster)
    meat <- meat + xu %*% t(xu)
  }

  cluster_count <- n_distinct(cluster)
  obs_count <- nrow(X)
  parameter_count <- ncol(X)
  correction <- (cluster_count / (cluster_count - 1)) *
    ((obs_count - 1) / (obs_count - parameter_count))

  correction * xtx_inv %*% meat %*% xtx_inv
}

tidy_clustered <- function(model, data, cluster_var) {
  clustered_vcov <- cluster_vcov(model, data[[cluster_var]])
  estimate <- coef(model)
  clustered_se <- sqrt(diag(clustered_vcov))
  model_rows <- suppressWarnings(as.integer(rownames(model.frame(model))))
  clusters <- if (all(!is.na(model_rows)) && max(model_rows) <= nrow(data)) {
    data[[cluster_var]][model_rows]
  } else {
    data[[cluster_var]][seq_len(nobs(model))]
  }
  cluster_count <- n_distinct(clusters)

  tibble(
    term = names(estimate),
    estimate = as.numeric(estimate),
    std.error = clustered_se,
    statistic = estimate / clustered_se,
    p.value = 2 * pt(abs(statistic), df = cluster_count - 1, lower.tail = FALSE)
  )
}

season_cluster_p_value <- function(model, data, term_name) {
  tidy_clustered(model, data, "season") |>
    filter(term == term_name) |>
    pull(p.value)
}

interaction_table <- bind_rows(
  tidy_clustered(model_tanking_index_interaction, candidate_data, "season") |>
    mutate(Outcome = "Tanking index")
) |>
  filter(term %in% c(
    "scale(top4_avg_rating)",
    "post_reformTRUE",
    "scale(top4_avg_rating):post_reformTRUE"
  )) |>
  mutate(
    term = recode(
      term,
      `scale(top4_avg_rating)` = "Top-four draft rating",
      post_reformTRUE = "Post-2019 reform",
      `scale(top4_avg_rating):post_reformTRUE` = "Top-four draft rating x post-2019"
    )
  ) |>
  select(
    Outcome,
    Term = term,
    Estimate = estimate,
    `Clustered std. error` = std.error,
    `Season-clustered p-value` = p.value
  )

interaction_terms <- tidy_clustered(model_tanking_index_interaction, candidate_data, "season")
interaction_pre_slope_est <- interaction_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(estimate)
interaction_pre_slope_p <- interaction_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(p.value)
interaction_change_est <- interaction_terms |>
  filter(term == "scale(top4_avg_rating):post_reformTRUE") |>
  pull(estimate)
interaction_change_p <- interaction_terms |>
  filter(term == "scale(top4_avg_rating):post_reformTRUE") |>
  pull(p.value)
interaction_post_slope_est <- interaction_pre_slope_est + interaction_change_est

model_table <- bind_rows(
  tidy(model_tanking_index) |>
    mutate(Model = "Tanking index")
) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = recode(
      term,
      `scale(marginal_lottery_incentive)` = "Position-adjusted draft incentive",
      pre_win_pct = "Pre-March win pct",
      season_2012TRUE = "2012 season",
      season_2020TRUE = "2020 season",
      season_2021TRUE = "2021 season"
    )
  ) |>
  select(
    Model,
    Term = term,
    Estimate = estimate,
    `Std. error` = std.error,
    `p-value` = p.value
  )

tanking_index_terms <- tidy(model_tanking_index)

tanking_index_incentive_est <- tanking_index_terms |> filter(term == "scale(marginal_lottery_incentive)") |> pull(estimate)
tanking_index_incentive_p <- tanking_index_terms |> filter(term == "scale(marginal_lottery_incentive)") |> pull(p.value)

top35_index_p <- tidy(model_tanking_index_top35) |>
  filter(term == "scale(marginal_lottery_incentive_top35)") |>
  pull(p.value)

control_correlation <- cor(
  candidate_data$pre_win_pct,
  candidate_data$games_back_cutline,
  use = "complete.obs"
)

tanking_index_no_2021_terms <- tidy(model_tanking_index_no_2021)

tanking_index_no_2021_incentive_est <- tanking_index_no_2021_terms |> filter(term == "scale(marginal_lottery_incentive)") |> pull(estimate)
tanking_index_no_2021_incentive_p <- tanking_index_no_2021_terms |> filter(term == "scale(marginal_lottery_incentive)") |> pull(p.value)

sensitivity_table <- bind_rows(
  tidy(model_tanking_index) |>
    mutate(Outcome = "Tanking index", Sample = "All years"),
  tidy(model_tanking_index_no_2021) |>
    mutate(Outcome = "Tanking index", Sample = "Excluding 2021")
) |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  mutate(
    term = "Position-adjusted draft incentive",
    Sample = factor(Sample, levels = c("All years", "Excluding 2021")),
    Outcome = factor(Outcome, levels = c("Tanking index"))
  ) |>
  arrange(Outcome, Sample, term) |>
  select(
    Outcome,
    Sample,
    Term = term,
    Estimate = estimate,
    `p-value` = p.value
  )

team_fit_correlation <- cor(
  candidate_data$team_fit_score,
  candidate_data$top4_avg_rating,
  use = "complete.obs"
)

team_fit_table <- bind_rows(
  tidy(model_tanking_index_fit) |>
    mutate(Outcome = "Tanking index")
) |>
  filter(term %in% c(
    "scale(top4_avg_rating)",
    "scale(team_fit_score)",
    "scale(top4_avg_rating):scale(team_fit_score)"
  )) |>
  mutate(
    term = recode(
      term,
      `scale(top4_avg_rating)` = "Top-four draft rating",
      `scale(team_fit_score)` = "Team fit",
      `scale(top4_avg_rating):scale(team_fit_score)` = "Top-four draft rating x team fit"
    )
  ) |>
  select(
    Outcome,
    Term = term,
    Estimate = estimate,
    `p-value` = p.value
  )

fit_index_interaction_p <- tidy(model_tanking_index_fit) |>
  filter(term == "scale(top4_avg_rating):scale(team_fit_score)") |>
  pull(p.value)

fit_index_interaction_est <- tidy(model_tanking_index_fit) |>
  filter(term == "scale(top4_avg_rating):scale(team_fit_score)") |>
  pull(estimate)

core_usage_table <- bind_rows(
  tidy(model_core_usage) |>
    mutate(Model = "Core-player usage drop rate")
) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = recode(
      term,
      `scale(top4_avg_rating)` = "Top-four draft rating",
      core_mpg_pre_march = "Pre-March core MPG",
      season_2012TRUE = "2012 season",
      season_2020TRUE = "2020 season",
      season_2021TRUE = "2021 season"
    )
  ) |>
  select(
    Model,
    Term = term,
    Estimate = estimate,
    `Std. error` = std.error,
    `p-value` = p.value
  )

core_usage_terms <- tidy(model_core_usage)
core_usage_incentive_est <- core_usage_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(estimate)
core_usage_incentive_p <- core_usage_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(p.value)
core_usage_model_n <- nobs(model_core_usage)
core_usage_incentive_pp <- 100 * core_usage_incentive_est
core_usage_incentive_abs_pp <- abs(core_usage_incentive_pp)
core_usage_direction <- if_else(core_usage_incentive_est >= 0, "larger", "smaller")
core_usage_incentive_cluster_p <- season_cluster_p_value(
  model_core_usage,
  candidate_data,
  "scale(top4_avg_rating)"
)
core_usage_cluster_count <- n_distinct(candidate_data$season[!is.na(candidate_data$core_usage_drop_rate)])

extract_core_usage_term <- function(model, data, model_label, term_name, term_label) {
  tidy(model) |>
    filter(term == term_name) |>
    transmute(
      Model = model_label,
      Term = term_label,
      `Estimate (pp)` = 100 * estimate,
      `p-value` = p.value,
      `Season-clustered p-value` = season_cluster_p_value(model, data, term_name),
      N = nobs(model),
      `Adj. R2` = summary(model)$adj.r.squared
    )
}

core_usage_sensitivity_table <- bind_rows(
  extract_core_usage_term(
    model_core_usage,
    candidate_data,
    "Raw top-four rating",
    "scale(top4_avg_rating)",
    "Top-four draft rating"
  ),
  extract_core_usage_term(
    model_core_usage_interaction,
    candidate_data,
    "Pre/post interaction",
    "scale(top4_avg_rating)",
    "Top-four rating"
  ),
  extract_core_usage_term(
    model_core_usage_interaction,
    candidate_data,
    "Pre/post interaction",
    "scale(top4_avg_rating):post_reformTRUE",
    "Top-four rating x post-2019"
  ),
  extract_core_usage_term(
    model_core_usage_position_adjusted,
    candidate_data,
    "Position-adjusted incentive",
    "scale(marginal_lottery_incentive)",
    "Position-adjusted incentive"
  ),
  extract_core_usage_term(
    model_core_usage_position_adjusted_no_2021,
    candidate_data_no_2021,
    "Position-adjusted, excluding 2021",
    "scale(marginal_lottery_incentive)",
    "Position-adjusted incentive"
  ),
  extract_core_usage_term(
    model_core_usage_no_2021,
    candidate_data_no_2021,
    "Raw top-four, excluding 2021",
    "scale(top4_avg_rating)",
    "Top-four draft rating"
  ),
  extract_core_usage_term(
    model_core_usage_fit,
    candidate_data,
    "Team-fit interaction",
    "scale(top4_avg_rating):scale(team_fit_score)",
    "Top-four rating x team fit"
  ),
  extract_core_usage_term(
    model_core_usage_top35,
    candidate_data,
    "Top-35 rating",
    "scale(avg_rating)",
    "Top-35 draft rating"
  )
)

kable(
  collapse_repeated_rows(interaction_table, "Outcome"),
  digits = 3,
  caption = "Table 2. Exploratory pre/post lottery reform interaction model with standard errors clustered by season."
)
```

The interaction models were included because feedback suggested testing whether the relationship between draft strength and tanking signals changed after the 2019 lottery reform. They are useful as an exploratory check, but they are not our preferred final model. With only 17 draft-class seasons, splitting the draft-strength effect into pre- and post-reform components makes the model harder to estimate and easier for individual seasons to influence.

The coefficient of interest in Table 2 is the draft-strength slope and its interaction with the post-2019 reform indicator. Because `top4_avg_rating` is standardized, the top-four draft rating coefficient means that, before the reform, a one-standard-deviation increase in top-four draft strength is associated with a `r round(interaction_pre_slope_est, 3)` change in the tanking index, holding pre-March strength and disrupted-season controls fixed. The interaction coefficient of `r round(interaction_change_est, 3)` is the additional post-reform change in that slope, so the implied post-reform slope is `r round(interaction_post_slope_est, 3)`. Since draft strength is shared by teams in the same season, Table 2 reports standard errors clustered by season. The season-clustered interaction p-value is `r format.pval(interaction_change_p, digits = 2)`, so we do not treat this as strong evidence that the relationship changed after the lottery reform.

### Analysis 2: Position-Adjusted Team-Season Regression

For that reason, our final model pools all seasons together and focuses on the team-specific position-adjusted draft incentive. This keeps the model simpler than the pre/post interaction while addressing an important weakness of the raw draft-rating model: not every likely lottery team has the same chance of turning a strong top-four draft class into a top-four pick.

We considered including both `pre_win_pct` and `games_back_cutline` as model controls, but they are highly correlated among likely lottery candidates, with a correlation of `r round(control_correlation, 3)`. To avoid unnecessary multicollinearity, the final model keeps `pre_win_pct` as the main baseline team-quality control and removes `games_back_cutline` from the regression. `games_back_cutline` still matters in the analysis because it helps define which teams are likely lottery candidates, but it is not included as an additional predictor in the final model.

We also compared two ways of measuring draft strength: the average rating of the top 35 prospects and the average rating of only the top four prospects. The top-four version is our preferred measure because the lottery is mainly about access to the top of the draft. In both models, the coefficient of interest has the same interpretation: the expected change in `tanking_index` for a one-standard-deviation increase in the position-adjusted draft incentive. The top-four version performs slightly better in the final model: the top-35 incentive p-value is `r format.pval(top35_index_p, digits = 2)`, compared with `r format.pval(tanking_index_incentive_p, digits = 2)` for the top-four incentive.

The final model uses the full `tanking_index` rather than a binary collapse flag because the index keeps more information. A strict cutoff is useful for labeling the most obvious cases in figures, but it treats a small late-season decline and a strong late-season improvement as equally "not flagged." The continuous index is better for regression because larger underperformance counts more strongly, while late-season overperformance counts against the tanking-incentive story.

```{r final-model-table}
kable(
  collapse_repeated_rows(model_table, "Model"),
  digits = 3,
  caption = "Table 3. Final pooled regression models using position-adjusted draft incentives among likely lottery candidates."
)
```

The tanking-index model has a positive position-adjusted incentive coefficient of `r round(tanking_index_incentive_est, 3)`, with a p-value of `r format.pval(tanking_index_incentive_p, digits = 2)`. Because the incentive variable is standardized, this means that a likely lottery team with a one-standard-deviation larger marginal top-four draft incentive is expected to have a tanking index that is `r round(tanking_index_incentive_est, 3)` units higher, holding pre-March win percentage and disrupted-season controls fixed. Positive values of the tanking index mean more late-season wins below expectation, so the positive coefficient points in the direction predicted by the tanking-incentive story.

Because 2021 is visibly unusual in the exploratory graphs, we also refit the same all-years models after removing only the 2021 season. This is a sensitivity check, not a replacement for the main model. The main model uses all available years from 2010 through 2026, while the sensitivity model asks whether the conclusion depends heavily on one COVID-affected season.

```{r no-2021-sensitivity}
kable(
  collapse_repeated_rows(sensitivity_table, "Outcome"),
  digits = 3,
  caption = "Table 4. Sensitivity check comparing position-adjusted models with models excluding 2021."
)
```

Excluding 2021 does not reverse the direction of the estimate. The coefficient has the same interpretation as in Table 3, but it is estimated on the sample without the unusual 2021 season: a one-standard-deviation larger position-adjusted draft incentive is associated with a `r round(tanking_index_no_2021_incentive_est, 3)` higher tanking index, holding the remaining controls fixed. The position-adjusted incentive coefficient changes from `r round(tanking_index_incentive_est, 3)` in the all-years model to `r round(tanking_index_no_2021_incentive_est, 3)` after excluding 2021, with the p-value changing from `r format.pval(tanking_index_incentive_p, digits = 2)` to `r format.pval(tanking_index_no_2021_incentive_p, digits = 2)`. This suggests that 2021 matters, but the main pattern is not simply a one-season artifact. The separate season controls in the main model let us keep all years in the primary analysis while acknowledging that disrupted seasons are not perfectly comparable to normal 82-game seasons.

### Analysis 3: Team-Fit Heterogeneity Check

One peer-review suggestion was to account for whether a draft class fits a particular team's needs. This is a useful idea because a strong draft class may not create the same incentive for every team. For example, a team that badly needs guard creation may respond differently to a guard-heavy draft than to a draft whose best prospects are mostly bigs.

We build a rough `team_fit_score` by matching team need to draft supply. Team need comes from pre-March production by position group. For guards, wings, and bigs, we calculate each team's `value_per36` using points, rebounds, assists, steals, blocks, turnovers, and minutes. We then standardize that rate within each season-position group, so below-average production becomes a positive need score and above-average production becomes 0 need. Draft supply comes from the top 35 CraftedNBA prospects: we calculate each prospect's positive rating value above 50, sum that value by position group, and convert it to a share of the draft class's top-35 value. The final score multiplies each position group's team need by that position group's draft-value share and sums across guards, wings, and bigs.

This extension is exploratory, so it is not our preferred final model. It is still useful because `team_fit_score` is team-specific and is only weakly correlated with top-four draft rating among likely lottery candidates (`r round(team_fit_correlation, 3)`). In other words, it is not just another version of draft strength.

```{r team-fit-visual, fig.width=8, fig.height=5.4, fig.cap="Figure 6. Team fit and tanking index among likely lottery candidates. Each point is one team-season. The x-axis measures whether a team's pre-March positional weaknesses align with the strongest prospect position groups in that year's draft, and color shows top-four draft class strength."}
candidate_data |>
  select(
    season,
    team_abbreviation,
    top4_avg_rating,
    team_fit_score,
    tanking_index
  ) |>
  ggplot(aes(x = team_fit_score, y = tanking_index, color = top4_avg_rating)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(alpha = 0.75, size = 2.3) +
  geom_smooth(method = "lm", se = FALSE, color = "gray35", linewidth = 0.8) +
  scale_color_gradient(
    low = "#5D6D7E",
    high = "#B23A48",
    name = "Top-four rating"
  ) +
  labs(
    title = "Team fit is a weak exploratory signal",
    x = "Team fit score",
    y = "Tanking index"
  )
```

```{r team-fit-table}
kable(
  collapse_repeated_rows(team_fit_table, "Outcome"),
  digits = 3,
  caption = "Table 5. Exploratory models adding team fit and a draft-strength-by-fit interaction."
)
```

The coefficient of interest in Table 5 is the interaction between top-four draft strength and team fit. Because both variables are standardized, the interaction coefficient of `r round(fit_index_interaction_est, 3)` means that a one-standard-deviation increase in team fit changes the draft-strength slope by `r round(fit_index_interaction_est, 3)` tanking-index units, holding the controls fixed. The interaction p-value is `r format.pval(fit_index_interaction_p, digits = 2)`, so this is not statistically significant. This does not mean fit is irrelevant; it means our simple fit proxy does not add clear evidence beyond the main draft-strength model. A stronger version of this idea would use team-specific draft boards, roster contract data, or play-type needs instead of broad position groups.

### Analysis 4: Core-Player Usage Reduction

The final exploratory model uses a different kind of tanking signal: whether likely lottery teams reduce the minutes of their main pre-March players late in the season. This is motivated by the idea that tanking may show up not only in losses, but also in roster management. We define each team's core players as the three players with the most pre-March starts, breaking ties by pre-March minutes. To focus on usage changes rather than trades or complete absences, we keep only core players who appeared for the same team at least once after March 1. We then calculate their combined minutes per team game before March and after March. The outcome, `core_usage_drop_rate`, is the percentage drop in those core-player minutes after March 1. Positive values mean the team's main continuing pre-March players were used less late in the season.

This model should be interpreted cautiously. The player logs include `did_not_play`, `active`, and `reason` fields, but they do not perfectly tell us whether a player was healthy. Therefore this is a **core-player usage reduction** model, not proof that teams intentionally benched healthy starters. We use raw top-four draft rating here, instead of the position-adjusted lottery incentive, because this mechanism asks whether stronger top-end classes are associated with a broader roster-management response among likely lottery teams.

```{r core-usage-visual, fig.width=8, fig.height=5.4, fig.cap="Figure 7. Top-four draft strength and core-player usage reduction among likely lottery candidates. Each point is one team-season. Core players are the top three pre-March starters, among players who stayed with the same team and appeared after March 1. The y-axis is the percentage drop in their combined minutes per team game after March 1."}
candidate_data |>
  filter(!is.na(core_usage_drop_rate)) |>
  mutate(label = if_else(core_usage_drop_rate >= quantile(core_usage_drop_rate, 0.95, na.rm = TRUE),
                         paste(team_abbreviation, season),
                         NA_character_)) |>
  ggplot(aes(x = top4_avg_rating, y = core_usage_drop_rate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_jitter(aes(color = season), width = 0.12, height = 0, size = 2.2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "gray35", linewidth = 0.8) +
  geom_text_repel(
    aes(label = label),
    na.rm = TRUE,
    size = 2.7,
    seed = 470,
    box.padding = 0.35,
    min.segment.length = 0
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_viridis_c(option = "plasma", breaks = seq(2010, 2026, by = 4)) +
  labs(
    title = "Core-player usage drops are larger in stronger top-four classes",
    x = "CraftedNBA average top-four prospect rating",
    y = "Core-player usage drop after March 1",
    color = "Season"
  )
```

```{r core-usage-table}
kable(
  collapse_repeated_rows(core_usage_table, "Model"),
  digits = 3,
  caption = "Table 6. Exploratory model predicting late-season core-player usage reduction among likely lottery candidates."
)
```

The coefficient of interest in Table 6 is the standardized top-four draft rating. Because the draft rating is standardized and the outcome is a percentage drop in core-player minutes, the coefficient of `r round(core_usage_incentive_est, 3)` means that a one-standard-deviation stronger top-four draft class is associated with a `r round(core_usage_incentive_abs_pp, 1)` percentage point `r core_usage_direction` drop in late-season core-player usage, holding pre-March core-player minutes and disrupted-season controls fixed. The usual team-season p-value is `r format.pval(core_usage_incentive_p, digits = 2)`, and the p-value remains `r format.pval(core_usage_incentive_cluster_p, digits = 2)` when standard errors are clustered by draft season. This matters because `top4_avg_rating` has only `r core_usage_cluster_count` unique season-level values even though the outcome is measured for many team-seasons. The result therefore gives evidence of a roster-management pattern associated with stronger draft classes, but it should not be interpreted as causal proof of tanking because injuries, planned rest, player development, and unobserved availability can also affect late-season minutes.

Because this result is stronger than the main win-loss models, we also re-used the earlier model structures with `core_usage_drop_rate` as the outcome. Table 7 reports both the usual team-season p-value and a season-clustered p-value. The raw top-four draft rating remains positive when 2021 is excluded, but the position-adjusted lottery incentive does not explain the core-usage outcome. This suggests that core-player minutes respond more to the overall strength of the top of the draft than to the exact marginal lottery-odds gain from one March lottery position. The pre/post interaction is also not strong, so we do not treat the 2019 lottery reform as clearly changing the core-usage relationship. The core-usage team-fit interaction is interesting under clustered inference, but because it is a secondary interaction built from a rough fit proxy, we treat it as hypothesis-generating rather than as a main result.

```{r core-usage-sensitivity-table}
kable(
  collapse_repeated_rows(core_usage_sensitivity_table, "Model"),
  digits = 3,
  caption = "Table 7. Sensitivity checks using core-player usage reduction as the outcome."
)
```

### Conclusions From the Formal Analysis

The main finding is that draft strength appears related to late-season losing signals, but the strength of the evidence depends on how directly we model incentives. The raw draft-strength figures show a clearer positive relationship, but they are partly season-level summaries. The position-adjusted model is more conservative because it asks whether a team's own lottery position gives it more to gain from a strong class. In that model, the estimate remains positive for the tanking index, but it is not strong enough to support a causal claim.

The analysis also shows why the definition of tanking matters. Instead of forcing teams into a yes/no tanking category, the full tanking index measures degree of underperformance. That is better suited for regression because a small late-season decline, a major collapse, and a late-season improvement should not all be treated the same.

The team-fit extension adds a second check on heterogeneity. It asks whether a strong class matters more when its strongest prospect positions line up with a team's roster weaknesses. In our version, the answer is not clear. The fit interaction is not statistically significant, so we do not make it part of the main conclusion. Still, including the extension makes the model more responsive to the basketball mechanism: teams are not only responding to abstract draft strength, but to the players they could realistically hope to add.

The core-player usage model adds a different type of outcome. It asks whether stronger top-four draft classes are associated with using main continuing pre-March players less often late in the season. This is closer to a behavioral tanking mechanism than the win-loss outcome alone, and this version points in the expected direction: stronger top-four classes are associated with larger percentage drops in core-player minutes. However, the raw top-four rating is season-level, so we interpret this as a draft-class-level signal with team-season outcomes rather than as 172 independent draft-strength observations. The sensitivity checks show that this result is stronger for raw draft strength than for the position-adjusted lottery incentive. That makes the result interesting, but still exploratory: injuries, planned rest, player development, and unobserved availability can all reduce minutes for reasons unrelated to intentional tanking.

The 2021 season remains an important case. Several teams underperformed sharply late in the season, but that year was shortened and affected by COVID-era scheduling and player availability. Rather than deleting 2021 from the main model, we include a separate 2021 control. We also include separate controls for 2012 and 2020 because those seasons were disrupted in different ways. This keeps the observations in the main analysis while acknowledging that they may not be comparable to normal 82-game seasons.

## VI. Next Steps

The results should be interpreted cautiously. The model finds a positive relationship between position-adjusted top-four draft incentives and late-season underperformance, but it does not prove intent and should not be read as definitive evidence that teams deliberately lost games.

### Limitations and Cautions

The biggest limitation is that tanking is not directly observable. We can observe losses, late-season underperformance, and playoff context, but we cannot observe whether a front office intended to lose. For that reason, all outcomes in this report should be interpreted as tanking proxies or tanking signals.

A second limitation is that raw draft strength varies by season, not by team. Even after expanding the data to 2010-2026, there are only 17 draft-class observations. The position-adjusted incentive improves this by combining draft strength with each team's lottery position, but part of the signal still comes from season-level variation in draft quality. The incentive measure is also simplified: it uses the marginal gain in top-four-pick probability from moving one lottery slot worse, so it does not distinguish between the value of the first, second, third, and fourth picks, larger multi-slot changes, player-specific team preferences, or the possibility that teams care about protecting an already-strong lottery position.

### Future Extensions

A direct next step is to improve the core-player usage model with better availability data. The version in this report uses player game logs to identify each team's main pre-March players who stayed with the team and appeared after March 1, then measures whether their minutes drop late in the season. This keeps the analysis focused on continuing players, but it still cannot perfectly distinguish injuries, rest, player development, and coach's decisions. A stronger version would merge injury reports, transaction data, and inactive-list reasons to separate strategic late-season roster management from ordinary player availability.

Another extension would improve the team-fit score. The current version uses broad position groups and pre-March box-score production, not detailed roster needs, contracts, player age, front-office preferences, or team-specific draft evaluations. The fit extension should therefore be interpreted as a first-pass robustness check rather than a final answer about roster construction.

A final extension would compare multiple draft-quality measures. The CraftedNBA rating is useful because it gives one consistent external measure of draft class quality, but it is still one source's model-based ranking. Future versions should test whether results change when draft strength is measured by top-prospect rating, number of elite prospects, consensus mock draft rankings, or later NBA outcomes such as win shares.

### Final Takeaway

This project finds suggestive evidence that draft class strength is related to late-season losing signals among likely lottery candidates. The clearest descriptive patterns come from season-level draft-strength graphs, while the more conservative position-adjusted win-loss model shows positive but noisier estimates. The core-player usage model adds a stronger exploratory signal that teams in better top-four draft classes reduced minutes more for continuing core players. The full tanking index is a better final win-loss outcome because it measures the size and direction of late-season underperformance rather than depending on an arbitrary cutoff.

The most defensible conclusion is therefore cautious: draft class quality appears relevant to tanking incentives, but the evidence is not strong enough to claim that stronger draft classes directly cause tanking. The project improves on simpler tanking analyses by combining playoff context, lottery position, standardized late-season underperformance, and draft class quality into one framework.

The exploratory team-fit extension for the main win-loss outcome is less conclusive. It gives a plausible way to think about team-level heterogeneity, but the broad fit proxy does not add strong statistical evidence beyond overall draft strength in the primary tanking-index model. That leaves team fit as an important direction for future work rather than a core result of this report.

## References

-   CraftedNBA. [Draft Class Rankings](https://craftednba.com/draft/class-rankings).
-   CraftedNBA. [Best NBA Draft Prospects Since 2010](https://craftednba.com/draft/best-prospects).
-   hoopR. [SportsDataverse hoopR package](https://hoopr.sportsdataverse.org/).
-   NBA.com. [NBA Board of Governors approves changes to Draft Lottery system](https://www.nba.com/news/nba-board-governors-approves-changes-draft-lottery-system).
-   NBA.com. [NBA Draft Lottery explainer](https://www.nba.com/news/nba-draft-lottery-explainer?lv=true).
-   NBA.com. [Ties broken for order of selection in NBA Draft 2018](https://www.nba.com/news/ties-broken-order-selection-nba-draft-2018).
-   Price, J., Soebbing, B. P., Berri, D., & Humphreys, B. R. (2010). [Tournament Incentives, League Policy, and NBA Team Performance Revisited](https://journals.sagepub.com/doi/10.1177/1527002510363103).
-   Taylor, B. A., & Trogdon, J. G. (2002). [Losing to Win: Tournament Incentives in the National Basketball Association](https://ideas.repec.org/a/ucp/jlabec/v20y2002i1p23-41.html).
