---
title: "Draft Class Strength and NBA Tanking Incentives"
subtitle: "Final project presentation"
author: "Nico L, Leo K, Colin C"
format:
  revealjs:
    theme: default
    css: final-presentation.css
    slide-number: true
    transition: fade
    incremental: false
    embed-resources: true
    html-math-method: plain
    width: 1280
    height: 720
    margin: 0.075
    chalkboard: false
    preview-links: auto
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r setup}
#| include: false
library(tidyverse)
library(lubridate)
library(broom)
library(ggrepel)
library(scales)
library(knitr)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", color = "#163B45", size = 15),
      plot.subtitle = element_text(color = "#52656F", size = 10.5),
      axis.title = element_text(color = "#1D252C", size = 10.5),
      axis.text = element_text(color = "#52656F", size = 9),
      panel.grid.minor = element_blank(),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8.5),
      plot.background = element_rect(fill = "#F7F6F0", color = NA),
      panel.background = element_rect(fill = "#F7F6F0", color = NA)
    )
)

team_box <- readRDS("data/nba_team_box_2010_2026.rds")
player_box <- readRDS("data/nba_player_box_2010_2026.rds")
draft_prospects <- read_csv("data/crafted_prospects_2010_2026.csv", show_col_types = FALSE)
draft_strength <- read_csv("data/crafted_draft_strength_2010_2026.csv", show_col_types = FALSE)

collapse_repeated_rows <- function(data, columns) {
  for (column in columns) {
    values <- as.character(data[[column]])
    repeated <- values == lag(values, default = NA_character_)
    values[replace_na(repeated, FALSE)] <- ""
    data[[column]] <- values
  }

  data
}

position_group_from_label <- function(position, archetype = rep(NA_character_, length(position))) {
  case_when(
    position %in% c("PG", "SG", "G") ~ "Guard",
    position %in% c("SF", "F") ~ "Wing",
    position %in% c("PF", "C") ~ "Big",
    archetype == "Guard" ~ "Guard",
    archetype %in% c("Wing", "Forward") ~ "Wing",
    archetype == "Big" ~ "Big",
    TRUE ~ NA_character_
  )
}
```

```{r data-build}
#| include: false
nba_teams <- c(
  "ATL", "BOS", "BKN", "NJ", "CHA", "CHI", "CLE", "DAL", "DEN", "DET",
  "GS", "HOU", "IND", "LAC", "LAL", "MEM", "MIA", "MIL", "MIN", "NO",
  "NY", "OKC", "ORL", "PHI", "PHX", "POR", "SA", "SAC", "TOR", "UTAH",
  "WSH"
)

conference_lookup <- tibble(
  team_abbreviation = nba_teams,
  conference = c(
    "East", "East", "East", "East", "East", "East", "East", "West", "West", "East",
    "West", "West", "East", "West", "West", "West", "East", "East", "West", "West",
    "East", "West", "East", "East", "West", "West", "West", "West", "East", "West",
    "East"
  )
)

regular_games <- team_box |>
  filter(
    season_type == 2,
    team_abbreviation %in% nba_teams,
    season %in% 2010:2026
  ) |>
  mutate(
    win = team_score > opponent_team_score,
    cutoff_date = as.Date(paste0(season, "-03-01")),
    season_period = if_else(game_date < cutoff_date, "pre_march", "late")
  ) |>
  left_join(conference_lookup, by = "team_abbreviation")

team_periods <- regular_games |>
  group_by(season, team_abbreviation, conference, season_period) |>
  summarise(
    games = n(),
    wins = sum(win, na.rm = TRUE),
    losses = games - wins,
    point_diff = mean(team_score - opponent_team_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = season_period,
    values_from = c(games, wins, losses, point_diff),
    values_fill = 0
  )

compute_lottery_odds <- function(weights, draws) {
  seeds <- seq_along(weights)
  top_pick_odds <- numeric(length(weights))
  top4_pick_odds <- numeric(length(weights))

  add_draw_path <- function(remaining, drawn, path_probability) {
    if (length(drawn) == draws) {
      final_order <- c(drawn, setdiff(seeds, drawn))
      top_pick_odds[final_order[1]] <<- top_pick_odds[final_order[1]] + path_probability
      top4_pick_odds[final_order[1:4]] <<- top4_pick_odds[final_order[1:4]] + path_probability
      return(invisible(NULL))
    }

    total_weight <- sum(weights[remaining])
    for (seed in remaining) {
      add_draw_path(
        remaining = setdiff(remaining, seed),
        drawn = c(drawn, seed),
        path_probability = path_probability * weights[seed] / total_weight
      )
    }
  }

  add_draw_path(seeds, integer(), 1)

  tibble(
    lottery_seed_march = seeds,
    top_pick_odds = top_pick_odds,
    top4_pick_odds = top4_pick_odds
  )
}

lottery_odds <- bind_rows(
  compute_lottery_odds(
    weights = c(250, 199, 156, 119, 88, 63, 43, 28, 17, 11, 8, 7, 6, 5),
    draws = 3
  ) |>
    mutate(post_reform = FALSE),
  compute_lottery_odds(
    weights = c(140, 140, 140, 125, 105, 90, 75, 60, 45, 30, 20, 15, 10, 5),
    draws = 4
  ) |>
    mutate(post_reform = TRUE)
) |>
  group_by(post_reform) |>
  arrange(lottery_seed_march, .by_group = TRUE) |>
  mutate(
    marginal_top4_pick_gain = pmax(0, lag(top4_pick_odds) - top4_pick_odds),
    marginal_top4_pick_gain = replace_na(marginal_top4_pick_gain, 0),
    marginal_top4_pick_gain = if_else(marginal_top4_pick_gain < 1e-8, 0, marginal_top4_pick_gain)
  ) |>
  ungroup()

team_data <- team_periods |>
  mutate(
    pre_win_pct = wins_pre_march / games_pre_march,
    postseason_cutline = if_else(season >= 2021, 10L, 8L)
  ) |>
  group_by(season, conference) |>
  arrange(desc(pre_win_pct), desc(wins_pre_march), .by_group = TRUE) |>
  mutate(
    conference_rank_march = row_number(),
    cutline_wins = wins_pre_march[postseason_cutline[1]],
    cutline_losses = losses_pre_march[postseason_cutline[1]],
    games_back_cutline = pmax(
      0,
      ((cutline_wins - wins_pre_march) + (losses_pre_march - cutline_losses)) / 2
    )
  ) |>
  ungroup() |>
  group_by(season) |>
  arrange(pre_win_pct, wins_pre_march, point_diff_pre_march, .by_group = TRUE) |>
  mutate(
    lottery_position_march = row_number(),
    lottery_seed_march = pmin(lottery_position_march, 14L)
  ) |>
  ungroup() |>
  mutate(
    late_win_pct = wins_late / games_late,
    expected_late_wins = games_late * pre_win_pct,
    expected_sd = sqrt(pmax(games_late * pre_win_pct * (1 - pre_win_pct), 0.001)),
    late_win_shortfall = expected_late_wins - wins_late,
    tanking_index = late_win_shortfall / expected_sd,
    likely_lottery_candidate = conference_rank_march > postseason_cutline &
      (games_back_cutline >= 3 | pre_win_pct < 0.400),
    post_reform = season >= 2019,
    season_2012 = season == 2012,
    season_2020 = season == 2020,
    season_2021 = season == 2021,
    lottery_era = if_else(post_reform, "Post-reform (2019-2026)", "Pre-reform (2010-2018)")
  ) |>
  left_join(
    draft_strength |>
      select(
        draft_year,
        avg_rating,
        top_prospect_rating,
        top4_avg_rating,
        top5_avg_rating
      ),
    by = c("season" = "draft_year")
  ) |>
  left_join(
    lottery_odds,
    by = c("post_reform", "lottery_seed_march")
  ) |>
  mutate(
    marginal_lottery_incentive = top4_avg_rating * marginal_top4_pick_gain,
    marginal_lottery_incentive_top35 = avg_rating * marginal_top4_pick_gain
  )

player_usage <- player_box |>
  filter(
    season_type == 2,
    season %in% 2010:2026,
    team_abbreviation %in% nba_teams
  ) |>
  mutate(
    cutoff_date = as.Date(paste0(season, "-03-01")),
    season_period = if_else(game_date < cutoff_date, "pre_march", "late"),
    minutes_num = suppressWarnings(as.numeric(minutes)),
    minutes_played = replace_na(minutes_num, 0),
    starter_flag = starter %in% TRUE
  )

team_games <- regular_games |>
  distinct(season, team_abbreviation, game_id, game_date, season_period)

late_core_appearances <- player_usage |>
  filter(season_period == "late", minutes_played > 0) |>
  distinct(season, team_abbreviation, athlete_id)

core_players <- player_usage |>
  filter(season_period == "pre_march", minutes_played > 0) |>
  group_by(season, team_abbreviation, athlete_id, athlete_display_name) |>
  summarise(
    pre_starts = sum(starter_flag, na.rm = TRUE),
    pre_minutes = sum(minutes_played, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(season, team_abbreviation) |>
  arrange(desc(pre_starts), desc(pre_minutes), .by_group = TRUE) |>
  slice_head(n = 3) |>
  ungroup() |>
  semi_join(
    late_core_appearances,
    by = c("season", "team_abbreviation", "athlete_id")
  )

core_usage <- core_players |>
  select(season, team_abbreviation, athlete_id) |>
  inner_join(
    team_games,
    by = c("season", "team_abbreviation"),
    relationship = "many-to-many"
  ) |>
  left_join(
    player_usage |>
      select(season, team_abbreviation, game_id, athlete_id, minutes_played),
    by = c("season", "team_abbreviation", "game_id", "athlete_id")
  ) |>
  mutate(minutes_played = replace_na(minutes_played, 0)) |>
  group_by(season, team_abbreviation, season_period) |>
  summarise(
    core_minutes_per_game = sum(minutes_played, na.rm = TRUE) / n_distinct(game_id),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = season_period,
    values_from = core_minutes_per_game,
    names_prefix = "core_mpg_"
  ) |>
  mutate(
    core_usage_drop = core_mpg_pre_march - core_mpg_late,
    core_usage_drop_rate = if_else(
      core_mpg_pre_march > 0,
      core_usage_drop / core_mpg_pre_march,
      NA_real_
    )
  )

team_data <- team_data |>
  left_join(core_usage, by = c("season", "team_abbreviation"))

candidate_data <- team_data |>
  filter(likely_lottery_candidate)

season_summary <- candidate_data |>
  group_by(season, lottery_era) |>
  summarise(
    top4_avg_rating = first(top4_avg_rating),
    mean_tanking_index = mean(tanking_index, na.rm = TRUE),
    candidate_count = n(),
    .groups = "drop"
  )

model_tanking_index <- lm(
  tanking_index ~ scale(marginal_lottery_incentive) +
    pre_win_pct +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

candidate_data_no_2021 <- candidate_data |>
  filter(season != 2021)

model_tanking_index_no_2021 <- lm(
  tanking_index ~ scale(marginal_lottery_incentive) +
    pre_win_pct +
    season_2012 +
    season_2020,
  data = candidate_data_no_2021
)

model_core_usage <- lm(
  core_usage_drop_rate ~ scale(top4_avg_rating) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_core_usage_position_adjusted <- lm(
  core_usage_drop_rate ~ scale(marginal_lottery_incentive) +
    core_mpg_pre_march +
    season_2012 +
    season_2020 +
    season_2021,
  data = candidate_data
)

model_terms <- tidy(model_tanking_index)
tanking_index_incentive_est <- model_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(estimate)
tanking_index_incentive_p <- model_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(p.value)

no_2021_terms <- tidy(model_tanking_index_no_2021)
tanking_index_no_2021_est <- no_2021_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(estimate)
tanking_index_no_2021_p <- no_2021_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(p.value)

core_usage_terms <- tidy(model_core_usage)
core_usage_top4_est <- core_usage_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(estimate)
core_usage_top4_p <- core_usage_terms |>
  filter(term == "scale(top4_avg_rating)") |>
  pull(p.value)

core_usage_position_terms <- tidy(model_core_usage_position_adjusted)
core_usage_position_est <- core_usage_position_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(estimate)
core_usage_position_p <- core_usage_position_terms |>
  filter(term == "scale(marginal_lottery_incentive)") |>
  pull(p.value)

summary_counts <- list(
  team_seasons = nrow(team_data),
  seasons = n_distinct(team_data$season),
  candidate_seasons = nrow(candidate_data),
  candidate_share = nrow(candidate_data) / nrow(team_data),
  top4_min = min(team_data$top4_avg_rating, na.rm = TRUE),
  top4_max = max(team_data$top4_avg_rating, na.rm = TRUE)
)
```

## Why This Project Matters

::: callout
**Research question:** Are stronger NBA draft classes associated with more late-season underperformance among teams with weak playoff outlooks?
:::

-   Tanking debates usually focus on lottery odds.
-   This project adds the value of the actual draft class.
-   We measure **tanking signals**, not intent.

:::::: stat-grid
::: stat
[`r summary_counts$seasons`]{.value}[NBA seasons, 2010-2026]{.label}
:::

::: stat
[`r summary_counts$team_seasons`]{.value}[team-seasons in final data]{.label}
:::

::: stat
[`r percent(summary_counts$candidate_share, accuracy = 1)`]{.value}[classified as likely lottery candidates]{.label}
:::
::::::

::: notes
Nico opens. Keep this to about 75 seconds: question, why it matters, and the caution that we study signals rather than intent.
:::

## Talk Map

::::::: rubric-map
<div>

<strong>Nico</strong><br>Question, motivation, and data

</div>

<div>

<strong>Leo</strong><br>Tanking index and lottery incentive construction

</div>

<div>

<strong>Colin</strong><br>Results, interpretation, and limitations

</div>

<div>

<strong>All</strong><br>Q&A after the final takeaway

</div>
:::::::

::: callout
Target length: **12-14 minutes** for slides, leaving about **5 minutes** for questions.
:::

::: notes
This slide makes coordination visible for the rubric. Move quickly.
:::

## Data Sources

-   **NBA team game logs:** regular-season team box scores from `hoopR`.
-   **NBA player game logs:** used to estimate late-season changes in core-player minutes.
-   **CraftedNBA draft ratings:** external ratings for draft classes and prospects.

::: takeaway
Each observation is a **team-season**. Draft quality changes by season, while team context changes by team.
:::

<p class="source-note">

Sources: hoopR; CraftedNBA draft class rankings and prospect data.

</p>

::: notes
Nico. The main point is that the project combines basketball outcomes with an external draft-strength measure.
:::

## How We Define A Tanking Signal

We split every regular season at **March 1**.

::: takeaway
**Tanking index = (expected late wins - actual late wins) / expected late-win standard deviation**
:::

-   `pre_win_pct` estimates team strength before the final stretch.
-   Positive tanking index = fewer late wins than expected.
-   "Likely lottery candidates" are teams outside the postseason cutline and at least 3 games back or below .400 before March.

::: notes
Leo. Emphasize that this avoids calling every bad team a tanking team. We compare teams against their own earlier baseline.
:::

## What The Candidate Filter Captures

```{r summary-table}
summary_table <- team_data |>
  mutate(group = if_else(likely_lottery_candidate, "Likely lottery candidates", "Other teams")) |>
  group_by(group) |>
  summarise(
    N = n(),
    `Pre win pct` = mean(pre_win_pct, na.rm = TRUE),
    `Late win pct` = mean(late_win_pct, na.rm = TRUE),
    `Mean tanking index` = mean(tanking_index, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(group))

kable(summary_table, digits = 3, align = "lrrrr")
```

::: callout
The candidate filter identifies weaker teams, but most of them do **not** sharply collapse late.
:::

::: notes
Leo. The table supports the methods: the filter finds the teams where incentives plausibly exist, and the continuous index keeps the result from depending on a binary label.
:::

## Late-Season Performance Vs. Pre-March Baseline

```{r visual-expected-vs-actual}
#| fig-width: 9.4
#| fig-height: 5.15
#| out-width: 92%
label_data <- team_data |>
  filter(likely_lottery_candidate) |>
  slice_max(tanking_index, n = 10, with_ties = FALSE)

ggplot(team_data, aes(x = pre_win_pct, y = late_win_pct)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray55") +
  geom_point(
    aes(color = tanking_index, shape = likely_lottery_candidate),
    size = 2.25,
    alpha = 0.72
  ) +
  geom_text_repel(
    data = label_data,
    aes(label = paste(team_abbreviation, season)),
    size = 2.8,
    seed = 470,
    box.padding = 0.35,
    min.segment.length = 0
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_color_gradient2(
    low = "#2F5D62",
    mid = "gray76",
    high = "#A63D40",
    midpoint = 0
  ) +
  labs(
    title = "Most teams stay near their pre-March pace",
    subtitle = "Labels show the largest positive tanking-index seasons among likely lottery candidates",
    x = "Pre-March win percentage",
    y = "March-onward win percentage",
    color = "Tanking index",
    shape = "Likely lottery candidate"
  )
```

::: notes
Leo. Points below the line underperformed late. Redder points had larger standardized late-season win shortfalls.
:::

## Measuring The Draft Incentive

We use top-end draft quality because the lottery is mainly about access to the top of the draft.

::: takeaway
**Position-adjusted incentive = top-four prospect rating x marginal gain in top-four-pick odds**
:::

-   Top-four average rating ranges from **`r round(summary_counts$top4_min, 1)`** to **`r round(summary_counts$top4_max, 1)`** across drafts.
-   The same draft class creates different incentives for different lottery positions.
-   Teams with no extra top-four probability gain from moving one slot worse get an incentive value of 0.

::: notes
Leo. This is the bridge from a season-level draft variable to a team-season incentive variable.
:::

## Draft Strength And Late-Season Underperformance

```{r visual-draft-strength}
#| fig-width: 9.4
#| fig-height: 5.15
#| out-width: 92%
ggplot(season_summary, aes(x = top4_avg_rating, y = mean_tanking_index, color = lottery_era)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray55") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  geom_point(aes(size = candidate_count), alpha = 0.88) +
  geom_text_repel(aes(label = season), size = 2.8, seed = 470, min.segment.length = 0) +
  scale_color_manual(
    values = c(
      "Pre-reform (2010-2018)" = "#52656F",
      "Post-reform (2019-2026)" = "#A63D40"
    )
  ) +
  scale_size_continuous(range = c(2.8, 7.2), name = "Likely lottery\nteams") +
  labs(
    title = "Stronger top-four classes line up with higher average tanking signals",
    subtitle = "Each point is one draft season summarized across likely lottery candidates",
    x = "CraftedNBA average top-four prospect rating",
    y = "Mean tanking index among likely lottery candidates",
    color = "Lottery era"
  )
```

::: notes
Colin. This is descriptive, not causal. Mention that draft strength has only 17 season-level observations.
:::

## Team-Specific Incentives Are Noisier

```{r visual-position-adjusted-incentive}
#| fig-width: 9.4
#| fig-height: 5.15
#| out-width: 92%
figure_data <- candidate_data |>
  arrange(season, team_abbreviation) |>
  group_by(zero_incentive = marginal_lottery_incentive == 0) |>
  mutate(
    marginal_lottery_incentive_plot = if_else(
      zero_incentive,
      marginal_lottery_incentive + ((row_number() - 1) %% 7) * 0.025,
      marginal_lottery_incentive
    )
  ) |>
  ungroup()

ggplot(figure_data, aes(x = marginal_lottery_incentive, y = tanking_index, color = lottery_seed_march)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(
    aes(x = marginal_lottery_incentive_plot),
    alpha = 0.72,
    size = 2.2
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "gray35", linewidth = 0.8) +
  scale_color_viridis_c(
    option = "viridis",
    direction = -1,
    breaks = c(1, 4, 7, 10, 14)
  ) +
  labs(
    title = "Team-specific incentive points in the expected direction",
    subtitle = "But late-season wins are still affected by injuries, trades, rest, and randomness",
    x = "Top-four rating x marginal top-four-pick probability gain",
    y = "Tanking index",
    color = "March lottery seed"
  )
```

::: notes
Colin. This is the conservative view: still positive, but much noisier than the season-average plot.
:::

## Main Regression Results

```{r model-results-table}
model_results <- tibble(
  Outcome = c(
    "Tanking index",
    "Tanking index",
    "Core-player usage drop"
  ),
  Model = c(
    "All seasons",
    "Excluding 2021",
    "All seasons"
  ),
  `Key predictor` = c(
    "Position-adjusted draft incentive",
    "Position-adjusted draft incentive",
    "Top-four draft rating"
  ),
  Estimate = c(
    tanking_index_incentive_est,
    tanking_index_no_2021_est,
    100 * core_usage_top4_est
  ),
  `p-value` = c(
    tanking_index_incentive_p,
    tanking_index_no_2021_p,
    core_usage_top4_p
  )
)

kable(
  model_results,
  digits = 3,
  align = "lllrr",
  caption = "Positive estimates mean stronger draft incentives are associated with larger tanking signals."
)
```

::: callout
The win-loss model is **suggestive**. The core-player usage model is a stronger exploratory signal.
:::

::: notes
Colin. Interpret the signs first. Do not oversell p-values. The 2021 sensitivity check is here because COVID-era disruptions matter.
:::

## Roster Management Signal

```{r core-usage-visual}
#| fig-width: 9.4
#| fig-height: 5.15
#| out-width: 92%
candidate_data |>
  filter(!is.na(core_usage_drop_rate)) |>
  mutate(
    label = if_else(
      core_usage_drop_rate >= quantile(core_usage_drop_rate, 0.95, na.rm = TRUE),
      paste(team_abbreviation, season),
      NA_character_
    )
  ) |>
  ggplot(aes(x = top4_avg_rating, y = core_usage_drop_rate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_jitter(aes(color = season), width = 0.12, height = 0, size = 2.2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "gray35", linewidth = 0.8) +
  geom_text_repel(
    aes(label = label),
    na.rm = TRUE,
    size = 2.7,
    seed = 470,
    box.padding = 0.35,
    min.segment.length = 0
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_viridis_c(option = "plasma", breaks = seq(2010, 2026, by = 4)) +
  labs(
    title = "Core-player usage drops are larger in stronger top-four classes",
    subtitle = "Outcome: late-season percentage drop in minutes for top three continuing pre-March starters",
    x = "CraftedNBA average top-four prospect rating",
    y = "Core-player usage drop after March 1",
    color = "Season"
  )
```

::: notes
Colin. Explain why this is closer to behavior than wins alone, then say it still cannot distinguish healthy rest from injuries or development minutes.
:::

## What We Learned

-   Draft strength appears relevant to NBA tanking incentives.
-   The cleanest win-loss result is positive but noisy.
-   Roster management may be a better follow-up signal than wins alone.
-   The project improves on simple tanking stories by combining playoff context, lottery position, and draft class quality.

::: takeaway
Most defensible conclusion: **stronger draft classes are associated with larger tanking signals, but the evidence does not prove intentional losing.**
:::

::: notes
Nico or Colin. This is the main interpretation slide. Land the cautious final claim clearly.
:::

## Limitations And Next Steps

-   Tanking intent is not directly observable.
-   Draft strength varies mostly by season, so there are only `r summary_counts$seasons` draft-class observations.
-   Disrupted seasons, especially 2012, 2020, and 2021, require caution.
-   Next improvements: injury reports, trade-deadline roster changes, strength of remaining schedule, and more draft-quality measures.

::: notes
All speakers can share this slide if needed. Keep it brief so the talk ends under 15 minutes.
:::

## Final Takeaway

::: callout
Stronger top-end draft classes do not automatically prove tanking, but they appear to make late-season losing signals more likely among teams already outside realistic playoff contention.
:::

**Questions?**

<p class="source-note">

References used in the full report: CraftedNBA, hoopR, NBA lottery rules, Taylor & Trogdon (2002), Price et al. (2010).

</p>

::: notes
Thank the audience and invite questions. Leave about five minutes.
:::
